Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-07-31 paper/manuscript.tex
references for introduction

%%%%%%%%%%%%%%%%%%%%
%  Introduction    %
%%%%%%%%%%%%%%%%%%%%
\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy (FE) calculations allow the calculation of energy differences associated with changes from one state to another, such as ligand binding to protein from solvent. The calculations use non-physical intermediate states where atoms are changed from to another or can be removed or inserted by decoupling or coupling with their environment, Figure 1. These so-called alchemical states permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, known as an absolute FE calculation, or modifying a substituent between two analogous ligands in a binding site, a relative FE calculation. Two common approaches are free energy perturbation (FEP) and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's (CITE). Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software (CITE). This foundation work, combined with the methodological, technological and hardware improvements of the last 5-10 years have led to an explosion of interest.  The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies (both relative~\cite{relative-binding-free-energies} and absolute~\cite{absolute-binding-free-energies}), ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), affinity changes due to mutations in therapeutic targets~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}. As the field of molecular simulation moves towards millisecond timescales accurate FE calculations on even more challenging problems will be in scope. In the meantime, today's user may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.

\todo[inline, color={green!20}]{GT: I've rewritten the intro in a more classic style, explain title and some later jargon, bit of background, why its important, what we are doing here. tried to keep brief.
FEP ref Zwanzig, R. W. High‐Temperature Equation of State by a Perturbation Method. I. Nonpolar Gases. J. Chem. Phys. 1954, 22, 1420-1426
TI ref Kirkwood, J. G. Statistical Mechanics of Fluid Mixtures. J. Chem. Phys. 1935, 3, 300-313.
early refs could include:
Jorgensen, W. L.; Ravimohan, C. Monte Carlo simulation of differences in free energies of hydration. J. Chem. Phys. 1985, 83, 3050-3054.
Kollman, P. Free energy calculations: applications to chemical and biochemical phenomena. Chem. Rev. 1993, 93, 2395-2417.
Wong, C. F.; McCammon, J. A. Dynamics and design of enzymes and inhibitors. J. Am. Chem. Soc. 1986, 108, 3830-3832.
Merz Jr, K. M.; Kollman, P. A. Free energy perturbation simulations of the inhibition of thermolysin: prediction of the free energy of binding of a new inhibitor. J. Am. Chem. Soc. 1989, 111, 5649-5658

}
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1/Fig1_dummy.pdf}
    \caption{{\bf What is alchemistry?}
    \todo[inline, color=green!20]{JDC: What is the concept for what this figure is supposed to communicate?}
    }
    \label{fig:fig1}
\end{figure}

@@

This best practices guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical free energy calculation? 
    \item How do I chose the right free energy protocol and run it? How do I analyze my simulations accurately? 
    \item What software tools are available to simulate alchemical free energy protocols? 
\end{itemize}
This guide is meant to serve as a comprehensive starting point for new practitioners, as well as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) to help with standard alchemical simulation and analysis practices. 
We will also give a brief overview of available software packages and their corresponding capabilities. 
The checklist we provide can serve as a good guide for a methods section both to the author of a free energy paper as well as a reviewer. 
\todo[inline, color={green!20}]{JDC: Do we only describe software that allows these best practices to be implemented in totality, or do we highlight when certain practices cannot be implemented in software we describe?}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth,angle=270]{paper/figures/fig2/Fig2.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:fig1}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

The concerns mentioned so far, such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup that is beyond the scope of this article, there are some additional points to consider. Formal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Check up-front because charge or tautomeric change would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? were certain co-factors or partner proteins required to augment activity? 
%

@@

In summary, the key aspects for successful drug discovery application are to work within the domain of applicability using high quality X-ray structure of the target compounds and testing the approach with a retrospective validation study to ensure the basics of system setup are understood and under control. Always assess the confidence in the predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. In a drug discovery setting it is inevitable to consider larger virtual libraries, in these cases again evaluate if shorter simulation time per  window can be sufficient for a ‘virtual screening’ mode of application. It is expected that future tools may allow efficient exploration of large virtual libraries. There are many accounts of success of alchemical FE calculations, the methods are undoubtedly showing good performance towards the goal of binding energy prediction. However, it is important to have realistic expectations. Structure based drug design projects are often capable of improving potency relatively quickly and the range of activity narrows to just two-to-three log units. It may seem hard to impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has been performed, see drug discovery strategy related discussions of Mobley Abel and co. From a drug discovery point of view, FEP is expanding its domain of applicability, there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.

>>>>

%%%%%%%%%%%%%%%%%%%%
%  Introduction    %
%%%%%%%%%%%%%%%%%%%%
\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy (FE) calculations allow the calculation of energy differences associated with changes from one state to another, such as ligand binding to protein from solvent~\cite{zwanzig1954hightemperature}. The calculations use non-physical intermediate states where atoms are changed from one to another or can be removed or inserted by decoupling or coupling with their environment, Figure 1. These so-called alchemical states permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, known as an absolute FE calculation, or modifying a substituent between two analogous ligands in a binding site, a relative FE calculation. Two common approaches are free energy perturbation (FEP) and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, biosimspace}. This foundation work, combined with the methodological, technological and hardware improvements of the last 5-10 years have led to an explosion of interest.  The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies (both relative~\cite{relative-binding-free-energies} and absolute~\cite{absolute-binding-free-energies}), ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), affinity changes due to mutations in therapeutic targets~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}. As the field of molecular simulation moves towards millisecond timescales accurate FE calculations on even more challenging problems will be in scope. In the meantime, today's user may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1/Fig1_dummy.pdf}
    \caption{{\bf What is alchemistry?}
    \todo[inline, color=green!20]{JDC: What is the concept for what this figure is supposed to communicate?}
    }
    \label{fig:fig1}
\end{figure}

@@

This best practices guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical free energy calculation? 
    \item How do I chose the right free energy protocol and run it? How do I analyze my simulations accurately? 
    \item What software tools are available to simulate alchemical free energy protocols? 
\end{itemize}
This guide is meant to serve as a comprehensive starting point for new practitioners, as well as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) to help with standard alchemical simulation and analysis practices. 
We will also give a brief overview of available software packages and their corresponding capabilities.
The checklist we provide can serve as a good guide for a methods section both to the author of a free energy paper as well as a reviewer. 
\todo[inline, color={green!20}]{JDC: Do we only describe software that allows these best practices to be implemented in totality, or do we highlight when certain practices cannot be implemented in software we describe? Asjsm: where this is known we should do this.}

@@

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig2/Fig2.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:fig1}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

The concerns mentioned so far, such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup that is beyond the scope of this article, there are some additional points to consider. Formal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Check up-front because charge or tautomeric change would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? Does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Were certain co-factors or partner proteins required to augment activity? 
%

@@

In summary, the key aspects for successful drug discovery application are to work within the domain of applicability using high quality X-ray structure of the target compounds and testing the approach with a retrospective validation study to ensure the basics of system setup are understood and under control. Always assess the confidence in the predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. In a drug discovery setting it is inevitable to consider larger virtual libraries, in these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application. It is expected that future tools may allow efficient exploration of large virtual libraries. There are many accounts of success of alchemical FE calculations, the methods are undoubtedly showing good performance towards the goal of binding energy prediction. However, it is important to have realistic expectations. Structure based drug design projects are often capable of improving potency relatively quickly and the range of activity narrows to just two-to-three log units. It may seem hard to impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has been performed, see drug discovery strategy related discussions of Mobley Abel and co. From a drug discovery point of view, FEP is expanding its domain of applicability, there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.
